### 2020-11-11 docs/manuscript.tex
progress on introduction

\PaperTitle{
    HAMFinder: Combining information to predict HLA-associated mutations with a
    Bayesian regression model
} % Article title

@@

\Keywords{human leucocyte anigen system, multiple sequence alignment, escape mutations, 
  Bayesian inference, sparsity, horseshoe, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
  % - what are HLA associated mutations and why identifying HLA escape is important
  % 
  % - current methods lack important aspects
  % - Progress in Bayesian methods are not reflected in current HLA models
  % - Bayesian statistics provides a coherent framework to combine information from different sources
  % - Using this kind of imformation and new methods improve predicted performance tremendously
  % -
  \\
  \textbf{Motivation} \\
  The human leucocyte antigen system (HLA) is of paramount importance to combat viral
  infections by presenting peptides on the cell surface via MHC I.
  In this way, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards
  virus variants that escape that immune recognition pathway, e.g. through point 
  mutations that decrease binding of the respective peptide to MHC I. \\

@@

>>>>

\PaperTitle{
    HAMDetector: Combining information to predict HLA-associated mutations with a
    Bayesian regression model
} % Article title

@@

\Keywords{human leucocyte antigen system, HLA, multiple sequence alignment, escape mutations, 
  viral escape, Bayesian inference, sparsity, horseshoe, epitope prediction} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
  \\
  \textbf{Motivation} \\
  The human leucocyte antigen system (HLA) is of paramount importance to combat viral
  infections by presenting peptides on the cell surface via MHC I.
  In this way, CD8+ cytotoxic T-Lymphocytes exert a strong selection pressure towards
  virus variants that escape that immune recognition pathway, e.g. through point 
  mutations that decrease binding of the respective peptide to MHC I. \\

@@

% - subset of peptidome presented on the cell surface via MHC I.
% - Cytotoxic t cells are selected to not recognize self.
% - by rule of exclusion, recognized peptides are foreign
% - t cells induce cytolytic activity and recruit other immune cells, reference

% - basic function of statistical analysis of sequence data
% - state of the art + brumme paper, phylogeny, linkage disequilibrium and codon covariation
% cite brumme and other paper

% - problems of p value based methods and statistical significance testing
% - highlight benefits of Bayesian approaches
\subsection{The HLA system}

One way how the human immune system is able to recognize intracellular viral infections
is through the human leucocyte antigen system:
In cells with active protein biosynthesis, proteins are continuously synthesized
and also degraded by a process called proteasomal degradation, which cleaves
proteins into linear peptides of varying length.
A small subset of these peptides is presented on the cell surface via
receptors called MHC class I. The genomic region encoding for MHC I is known to be
highly polymorphic, with more than 20000 different HLA alleles described today.
The resulting gene products differ in their binding properties, which means that
cells from different individuals present a highly diverse set of peptides on their
surface.
Cytotoxic T cells are selected during maturation to only weakly bind to 
peptide/MHC I complexes when the peptide originated from proteins of the usual proteome, 
but might be able to strongly bind to complexes of MHC I with peptides which are 
generated through proteosomal degradation of a viral protein.
Upon activation, T cells induce cytolytic activity and recruit other immune cells.

\subsection{HLA escape}
In this way, the HLA system exerts a strong selection pressure towards virus variants
that escape T cell recognition, for example through a point mutation
that results in reduced binding of an immunogenic peptide to MHC I or through a
set of mutations that alters the viral protein in such a way that it is cleaved into 
different peptides that are not recognized by the host's T cell repertoire.

The evolutionary events are complex and occur not only on the level of individuals, 
where a virus adapts to specific features of the host, but also on the population level,
because HLA alleles differ in their frequency across geographic region, as they are inherited
according to standard Mendelian rules.

Upon transmission, HLA escape mutations typically quickly revert to their wild type
as they usually reduce viral replicative capacity (if a mutation would increase
viral replicative capacity regardless of the presence of a given HLA allele, it would
probably already be the wild type). Kawashima et al
describe an escape mutation that is selected by HLA allele B\*53,
does not strongly affect viral replicative capacity and therefore slowly enriches
over time in Japan, where B\*53 commonly occurs.
How quickly a given escape mutation is selected upon transmission in a host depends on the
magnitude of the reduction in viral replicative capacity, on the strength of selection
pressure and also on the genetic background, e.g. some escape mutations require compensatory 
mutations which partly attenuate the negative impact on viral replicative capacity.

Studying HLA escape therefore provides the unique opportunity to gain insight into
viral evolution, on the host level but also on the population level.
Unfortunately, identifying HLA escape mutations is difficult in practice.

\subsection{Identifying HLA-escape mutations}
